### 2023-08-31 manuscript.tex
Addressed various wording clarifications as suggested by review comments

\section{Results}
\subsection{Pipeline overview}
Our pipeline follows a similar overall approach to other graph-based methods of orthology inference. First, protein sequences from annotated genomes are collected (Fig.~\ref{fig:graphical_overview}A), and homology searches are conducted between all query-target pairs of genomes (Fig.~\ref{fig:graphical_overview}B). The raw output from these homology searches is processed to yield best hits between pairs of sequences (Fig.~\ref{fig:graphical_overview}C). Next, the network of best hits is clustered into self-consistent orthologous groups (Fig.~\ref{fig:graphical_overview}D). Since genes can have multiplied associated isoforms, we then implemented a novel second clustering step where orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters with warm and cool colors, respectively (Fig.~\ref{fig:graphical_overview}E). Finally, representative sequences in each orthologous group are aligned (Fig.~\ref{fig:graphical_overview}F). In the following sections, we discuss each of these and other steps which were omitted for clarity in greater detail.

@@

\subsection{Selection of single copy orthologous groups}
The criteria for selecting orthologous groups for further analyses depends on the biological question under investigation. For example, studies of gene duplication will focus on orthologous groups with paralogs in some lineages but not in others. In contrast, analyses which assume functional conservation should restrict the orthologous groups to single copy orthologs since paralogs more frequently undergo functional divergence~\cite{Altenhoff2012, Pegueroles2013, Soria2014}. A simple method for identifying such groups is requiring each species to have exactly one associated gene. However, since the probability of at least one missing gene annotation approaches one as the total number of genomes increases, this is too restrictive and fails to leverage the redundancy of closely related genomes. Instead, a set of phylogenetic diversity criteria detailed in Table~\ref{stable:diversity_criteria} were applied to ensure the major lineages were represented in downstream analyses. Furthermore, genome-wide analyses should select one orthologous group per each of the previously identified gene groups as to not bias the results towards genes with many distinct groups of isoforms. In summary, orthologous groups failing the phylogenetic diversity criteria were first removed, and the representative for each gene group was chosen as the highest scoring orthologous group when ranked by the number species and the sum of the bit scores associated with each edge. This significantly reduced the number of orthologous groups from 22,813 to 8,566.

@@

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several hidden states~\cite{Felsenstein1996, Siepel2004}. Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each. Thus, we identified three distinct types of regions in the alignments, roughly corresponding to regions with strong phylogenetic evidence, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. For simplicity, however, we refer to the states that generate each type of region as 1, 2, and 3, respectively. To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left). The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree. The second is the annotation process which can erroneously exclude or include segments from a sequence. The result is an alignment of annotated sequences which contains evolutionary information obscured by noise from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence. To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols. The distributions for each state then consisted of two components derived from the encoded sequences (Fig.~\ref{fig:hmm}A, right). The first component models the gap pattern with a Markov process. This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process. These subprocess roughly correspond to changes caused by evolution and annotation, respectively. Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the ``gap stickiness'' as a beta-binomial random variable by counting the number of symbols that remain constant between columns. Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.

>>>>

\section{Results}
\subsection{Pipeline overview}
Our pipeline follows a similar overall approach to other graph-based methods of orthology inference. First, protein sequences from annotated genomes are collected (Fig.~\ref{fig:graphical_overview}A), and homology searches are conducted between all query-target pairs of genomes (Fig.~\ref{fig:graphical_overview}B). The raw output from these homology searches is processed to yield best hits between pairs of sequences (Fig.~\ref{fig:graphical_overview}C). Next, the network of best hits is clustered into self-consistent orthologous groups (Fig.~\ref{fig:graphical_overview}D). Since genes can have multiple associated isoforms, we then implemented a novel second clustering step where orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters in the top and bottom, respectively (Fig.~\ref{fig:graphical_overview}E). Finally, representative sequences in each orthologous group are aligned (Fig.~\ref{fig:graphical_overview}F). In the following sections, we discuss each of these and other steps which were omitted for clarity in greater detail.

@@

\subsection{Selection of single copy orthologous groups}
The criteria for selecting orthologous groups for further analyses depends on the biological question under investigation. For example, studies of gene duplication will focus on orthologous groups with paralogs in some lineages but not in others. In contrast, analyses which assume functional conservation should restrict the orthologous groups to single copy orthologs since paralogs more frequently undergo functional divergence~\cite{Altenhoff2012, Pegueroles2013, Soria2014}. A simple method for identifying such groups is requiring each species to have exactly one associated gene. However, since the probability of at least one missing gene annotation approaches one as the total number of genomes increases, this is too restrictive and fails to leverage the redundancy of closely related genomes. Instead, a set of phylogenetic diversity criteria detailed in Table~\ref{stable:diversity_criteria} were applied to ensure the major lineages were represented in downstream analyses. Briefly, these criteria group species into sets with the requirement that a minimum number of species in each set is present. Furthermore, genome-wide analyses should select one orthologous group per each of the previously identified gene groups as to not bias the results towards genes with many distinct groups of isoforms. In summary, orthologous groups failing the phylogenetic diversity criteria were first removed, and the representative for each gene group was chosen as the highest scoring orthologous group when ranked by the number species and the sum of the bit scores associated with each edge. This significantly reduced the number of orthologous groups from 22,813 to 8,566.

@@

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several hidden states~\cite{Felsenstein1996, Siepel2004}. Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each. Thus, we identified three distinct types of regions in the alignments, roughly corresponding to regions with strong phylogenetic evidence, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. For simplicity, however, we refer to the states that generate each type of region as 1, 2, and 3, respectively. To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left). The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree. The second is the annotation process which can erroneously exclude or include segments from a sequence. The result is an alignment of annotated sequences which contains evolutionary information obscured by noise from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence. To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols. The distributions for each state then consisted of two components derived from the encoded sequences (Fig.~\ref{fig:hmm}A, right). The first component models the gap pattern with a Markov process. This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process. These subprocesses roughly correspond to changes caused by evolution and annotation, respectively. Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the ``gap stickiness'' as a beta-binomial random variable by counting the number of symbols that remain constant between columns. Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.
